---
author: "Vasu Janjrukia, Sarah Larson, Jaehyun Park, Justin Wang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(lubridate)
library(kableExtra)
library(broman)
source("../scripts/viridis.R")
source("../scripts/ggprob.R")
theme_set(theme_minimal())
```

```{r, include=FALSE}

attendance_orig = read_csv("School_Attendance_by_Student_Group_and_District__2021-2022.csv",
                           name_repair = make.names) %>% 
  select(-Category, -Reporting.period, -Date.update, -District.code) %>% 
  rename(district_name = District.name,
         group = Student.group) %>% 
  pivot_longer(cols = -c(district_name, group), 
               names_pattern = "^X(\\d{4}\\.\\d{4})\\.(\\w+\\.\\w+).*$", 
               names_to = c("year", "name")) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  rename(count = student.count,
        attendance_rate = attendance.rate)

attendance_orig$year = attendance_orig$year %>% str_replace("\\.", "-")

attendance = attendance_orig %>% 
  filter(group == "Free/Reduced Price Meal Eligible" |
           group == "All Students") %>% 
  filter(year != "2021-2022") %>% 
  group_by(district_name) %>% 
  filter(!is.na(count) & !is.na(attendance_rate)) %>% 
  mutate(n = n()) %>%
  filter(n == 4) %>%
  select(-n) %>% 
  ungroup()
  
```


# Introduction

The COVID-19 pandemic had far reaching impacts on our daily lives. One area that was greatly affected was education. The main goal of our research was to examine if any groups of students were disproportionately affected during the pandemic. In particular, we will be examining attendance rates between students who qualified for free or reduced price lunch against the attendance rates of the student body as a whole. In this paper, we will show students eligible for free or reduced-price lunch experienced a distinct decline in attendance rates during the COVID-19 pandemic, highlighting the unequal impact on different socioeconomic groups.

# Background

This data set was collected by State of Connecticut Department of Education^[https://catalog.data.gov/dataset/school-attendance-by-student-group-and-district-2021-2022] to measure attendance rates for all of the school districts in the state. Data for certain groups were suppressed to either maintain anonymity or to prevent misinterpretation of statistics from very small sample sizes as being as representative as those from larger ones. As a result, schools that didn't have published statistics on the attendance of students who qualified for free/reduced price meals were excluded. In the state of Connecticut, students who qualify for free/ reduced price meals belong to households eligible for SNAP or TFA benefits^[https://portal.ct.gov/-/media/SDE/Nutrition/NSLP/Forms/FreeRed/Parent_Letter_FAQ_Free_Reduced_Meals_NSLP_SBP.docx], which mostly includes households at or under two times the Federal Poverty Level.

# Analysis

Below we can see the median attendance rates for students, grouped by year and whether they are eligible for free/reduced price meals.

```{r, echo = FALSE}
attendancetest1 = attendance %>%
  pivot_wider(names_from = group, 
              values_from = c(count, attendance_rate)) %>%
  mutate(normstudents = `count_All Students` - `count_Free/Reduced Price Meal Eligible`) %>%
  mutate(arnormstudents = (`count_All Students` * `attendance_rate_All Students` -
                             `count_Free/Reduced Price Meal Eligible` * `attendance_rate_Free/Reduced Price Meal Eligible`) / `normstudents`) %>%
  rename(`nonfree Count` = `normstudents`, 
         `nonfree AR` = `arnormstudents`, 
         `FRPME Count` = `count_Free/Reduced Price Meal Eligible`, 
         `FRPME AR` = `attendance_rate_Free/Reduced Price Meal Eligible`) %>%
  select(`district_name`, `year`, `nonfree Count`, `nonfree AR`, `FRPME Count`, `FRPME AR`) %>%
  pivot_longer(cols = c(`nonfree Count`, `FRPME Count`), 
               names_to = "Group", 
               values_to = "Count") %>%
  select(`district` = `district_name`, `Group`, `year`, `Count`, `nonfree AR`, `FRPME AR`) %>%
  mutate(Group = ifelse(Group == "nonfree Count", "nonfree", "FRPME"))

free1 = attendancetest1 %>%
  filter(Group == "FRPME") %>%
  select(!c(`nonfree AR`, `Count`)) %>%
  drop_na() %>%
  pivot_wider(names_from = `year`, values_from = `FRPME AR`) %>%
  mutate(difference = `2020-2021` - `2019-2020`)

nonfree1 = attendancetest1 %>%
  filter(Group == "nonfree") %>%
  select(!c(`FRPME AR`, `Count`)) %>%
  drop_na() %>%
  pivot_wider(names_from = `year`, values_from = `nonfree AR`) %>%
  mutate(difference = `2020-2021` - `2019-2020`) %>% 
  replace_na(list(`2019-2020` = 100, difference = 0))
```

```{r, echo = FALSE}

nonfree1_long = nonfree1 %>% 
  select(-difference) %>% 
  pivot_longer(cols = c(`2020-2021`, `2019-2020`),
               names_to = "year",
               values_to = "attendance_rate")

free1_long = free1 %>% 
  select(-difference) %>% 
  pivot_longer(cols = c(`2020-2021`, `2019-2020`),
               names_to = "year",
               values_to = "attendance_rate")


ggplot() +
  geom_density(data = free1_long, 
               aes(x = attendance_rate, 
                   fill = "Eligible for Free/Reduced Price Meals"), 
               alpha = 0.5) +
  geom_density(data = nonfree1_long, 
               aes(x = attendance_rate, 
                   fill = "Not Eligible for Free/Reduced Price Meals"), 
               alpha = 0.5) +
  facet_grid(rows = vars(year)) +
  labs(title = "Attendance rates for Students in Conn.",
       x = "Attendance Rate",
       y = "Density",
       fill = "Free/Reduced Price Meal Eligibility") + 
  theme(strip.text.y.right = element_text(angle = 0)) +
  scale_x_continuous(breaks = seq(0.75, 1, 0.1), limits = c(0.75, 1))

```

In the graph below, we can see the difference in attendance between students eligible for free/reduced price meals and students who are not eligible, broken up by school year. 

```{r, echo = FALSE}
  ggplot() +
    geom_density(data = free1, aes(x = difference, fill = "Eligible for Free/Reduced Price Lunch"), alpha = 0.5) +
    geom_density(data = nonfree1, aes(x = difference, fill = "Not Eligible for Free/Reduced Price Lunch"), alpha = 0.5) +
  labs(title = "Difference in Attendence between the 2019-2020 school year and 2020-2021 school year",
       x = "Difference between Attendance Rates",
       y = "Frequency",
       fill = "Eligibility")
```

We can see that means of the difference between students eligible for free/reduced meals and those who are not eligible, but in order to test if it is statistically significant, 

```{r, echo = FALSE}
t.test(nonfree1$difference, free1$difference)
```

# Discussion

We have sufficient evidence ($p$ = 2.025e-11) that the difference for attendance rates between 2019-2020 and 2020-2021 for free/reduced meal eligible students was greater than that for normal students. Thus, we can conclude that the pandemic negatively impacted free/reduced price meal eligible students more than those who are not eligible. Some possible short-comings are that we don’t have a general baseline before Covid happened because our baseline year (2019-2020) was still affected by Covid towards the end of the year. Another short-coming is that we don’t know how a school counted an absence, meaning that one school could count only missing a short amount of time an absence and another would only count an entire day missed as an absence. Future work could involve a more detailed exploration of these external factors and their impact on attendance disparities. Additionally, a deeper dive into subgroup analyses within the "Free/Reduced Price Meal Eligible" category may provide more nuanced insights.

Some topics of further research can cover how Covid-19 affected the attendance of students of different races or if there is a specific group of students that make up a large percentage of the free/price reduced meal eligible group?

Rather than finding the difference between school years for each group then comparing them with the other, we can find the difference between groups for each school year, then comparing them to see whether that difference is greater between school years. The inclusion of data from more years enables us to track attendance rate trends over an extended period, providing a nuanced understanding of how attendance patterns have evolved during the ongoing impact of the COVID-19 pandemic. Our primary conclusion is that the pandemic negatively impacted free/reduced price meal eligible students more than those who are not eligible. Our primary evidence that supports these conclusion is a two sample t-test done between the attendances differences from 2019-2020 to 2020-2021 of free/reduced price meal eligible students and those who are not eligible. The results of the t-test were t = 6.941 and a p-value of 2.025e-11

# References